\section{Methods}

Our approach operates in the latent space of pre-trained multilingual LLMs without fine-tuning. We identify language-specific directions via PCA on parallel translations, then steer representations along these directions to control code-switching.

\subsection{Language Direction Identification}

Given a parallel corpus $\mathcal{D} = \{(s_i, k_i)\}_{i=1}^N$ where $s_i$ represents semantically equivalent content in language $k \in \mathcal{K}$, we extract hidden states $\mathbf{h}_i^{(\ell)} \in \mathbb{R}^d$ from each layer $\ell$ of the LLM. Specifically, we use the hidden state at the last token position of each sentence. For each layer, we pool hidden states across all languages and apply PCA to the centered embeddings $\mathcal{H}^{(\ell)} = \{\mathbf{h}_i^{(\ell)}\}_{i=1}^N$. The first principal component $\mathbf{v}^{(\ell)}$ captures the \emph{language direction}:
\begin{equation}
  \mathbf{v}^{(\ell)} = \arg\max_{\|\mathbf{v}\|=1} \sum_{i=1}^N \left(\mathbf{v}^\top (\mathbf{h}_i^{(\ell)} - \bar{\mathbf{h}}^{(\ell)})\right)^2
\end{equation}
When semantic content is constant across languages, language identity becomes the dominant variance source, making $\mathbf{v}^{(\ell)}$ an interpretable language axis.

\subsection{Latent-Space Language Steering}

Empirical analysis (Section~\ref{sec:results}) shows language variance concentrates in final layers. We apply steering only to layers $\ell \geq \ell_{\text{crit}}$: for Qwen2.5-1.5B (28 layers), we steer layers 27-28; for Llama-3.2-1B (16 layers), we steer layers 15-16. For each token's hidden state $\mathbf{h}_t^{(\ell)}$ during generation, we modify the language component:
\begin{equation}
  \tilde{\mathbf{h}}_t^{(\ell)} = \mathbf{h}_t^{(\ell)} - s \left(\mathbf{h}_t^{(\ell)} \cdot \mathbf{v}^{(\ell)}\right) \mathbf{v}^{(\ell)}
\end{equation}
where $s \in \mathbb{R}$ controls intervention strength, selected via grid search to optimize either distributional similarity (KL divergence) or generation-based code-switching metrics (Section~\ref{sec:generation_eval}). When $s = 1$, this is an exact orthogonal projection that removes the language component. Values $0 < s < 1$ partially attenuate it, while $|s| > 1$ goes beyond projection removal, actively pushing representations away from the source language direction. In practice, optimal coefficients have $|s| > 1$ (see Table~\ref{tab:steering_coeff}), indicating that mere removal is insufficient and active counter-steering is needed to overcome the language bias in code-switched input.

\subsection{Language Prediction}

To validate our approach, we train a logistic regression classifier on final-layer projections $z_i = \mathbf{h}_i^{(L)} \cdot \mathbf{v}^{(L)}$:
\begin{equation}
  P(k \mid z) = \frac{\exp(w_k z + b_k)}{\sum_{k' \in \mathcal{K}} \exp(w_{k'} z + b_{k'})}
\end{equation}
This one-dimensional classifier predicts language identity with high accuracy, demonstrating the interpretability of discovered directions.
