\section{Conclusion}

We introduced latent-space language steering, a lightweight inference-time method that mitigates unintended code-switching in multilingual LLMs without fine-tuning. By extracting language directions via PCA on parallel translations, our approach enables precise control over language identity with negligible overhead—one dot product and vector subtraction per token.

Our empirical analysis across Qwen2.5 and Llama-3.2 models reveals that language identity concentrates in final layers with near-perfect linear separability. This geometric structure enables both accurate language classification (95-99\% accuracy) and effective steering that reduces distributional divergence by up to 55\% and achieves 63--99\% Code-Switching Index reduction in generated text. The layer-wise evolution shows language identity crystallizes in final layers while early layers focus on semantics, suggesting these occupy largely orthogonal subspaces.

While distributional divergence is reduced by 40\% on average and generated text shows near-complete language conversion (TLC $> 0.77$ across all pairs), steering coefficients are highly fragile - aggressive values eliminate code-switching but can produce degenerate repetition, revealing a sharp trade-off between language control and generation fluency. Our approach demonstrates that interpretable control mechanisms can be extracted from learned representations, offering a promising direction toward lightweight alternatives to fine-tuning for multilingual language control, though addressing the fluency--control trade-off and coefficient sensitivity remains important future work. Natural extensions include multilingual steering (controlling language choice among three or more languages simultaneously) and bidirectional steering (e.g., steering from English toward Spanish), both of which the geometric structure of the multilingual PCA maps suggests should be feasible.
