\chapter{Introduction}

\noindent\textit{\textbf{Note on the use of AI tools:} In some portions of this document (20\% - 30\% of the entire text), Artificial Intelligence assistants, particularly Generative AI, have been used to improve, rephrase, and summarize the content. The technologies used include Claude Code.}

\bigskip

Large language models (LLMs) have emerged as one of the most transformative technologies in artificial intelligence, demonstrating remarkable capabilities across diverse domains including natural language understanding, code generation, mathematical reasoning, and creative writing~\cite{brown2020language,openai2023gpt4,touvron2023llama}. These models, trained on vast corpora of text data, have shown an unprecedented ability to generalize across tasks and generate coherent, contextually appropriate responses. However, despite their impressive performance, the deployment of LLMs in real-world applications faces several fundamental challenges that limit their reliability, controllability, and efficiency.

\paragraph{Relevance.}
The widespread adoption of LLMs in critical applications---from healthcare diagnostics to legal document analysis, from educational tutoring to scientific research assistance---demands that these systems be trustworthy and predictable. Yet current LLMs exhibit three interconnected limitations that undermine their practical utility.

First, LLMs frequently produce incorrect or fabricated information with high apparent confidence, a phenomenon known as hallucination~\cite{ji2023survey}. Users cannot easily distinguish between reliable and unreliable model outputs, making it difficult to trust LLM-generated content without extensive verification. The gap between a model's confidence and its actual correctness poses critical risks, particularly in high-stakes domains where overconfidence in incorrect answers leads to severe consequences.

Second, once trained, LLMs exhibit fixed behavioral patterns that may not align with user requirements. For multilingual models, this manifests as unwanted code-switching, where models unexpectedly switch languages mid-response~\cite{winata2021language}. More broadly, fine-grained control over generation behavior remains challenging without expensive retraining, leaving operators with limited options when a model's default behavior does not match deployment requirements.

Third, fine-tuning LLMs on new tasks or domains requires substantial computational resources. Current approaches often treat all training examples equally, ignoring the varying complexity and informativeness of different data points, leading to wasted computation on redundant or trivial examples. This inefficiency is particularly problematic for organizations with limited computational budgets that nonetheless need to adapt models to specialized domains.

These challenges are not merely academic concerns---they represent practical barriers to LLM adoption in high-stakes applications where errors carry significant consequences.

\paragraph{Main purpose of the research.}
This thesis investigates a unified approach to addressing these three challenges by leveraging signals that are readily available from a single forward pass through the model. Specifically, we exploit two types of signals: (1) \textit{output distributions}, the probability distributions over tokens that models produce at each generation step, and (2) \textit{internal representations}, the high-dimensional activation patterns in the model's hidden layers. Our central hypothesis is that these signals encode rich information about model behavior, uncertainty, and input characteristics that can be extracted and utilized without requiring architectural modifications or additional training. Moreover, the interplay between these two signal types opens a path toward \textit{uncertainty decomposition}: the geometric structure of internal representations can separate language-specific from language-agnostic components, potentially enabling more fine-grained analysis of why a model is uncertain.

The research addresses three concrete questions.
\begin{enumerate}
  \item \textbf{RQ1}: How can patterns in token-level output distributions be used to characterize model uncertainty and predict answer correctness?
  \item \textbf{RQ2}: What geometric structure exists in the latent representations of multilingual LLMs, and how can this structure be exploited to control language-specific behaviors?
  \item \textbf{RQ3}: How can output distribution characteristics guide the selection and processing of training data to improve fine-tuning efficiency?
\end{enumerate}

\paragraph{Scientific novelty.}
This thesis makes several novel contributions to the field of LLM research.

We present a systematic analysis of token-wise entropy as an uncertainty indicator, demonstrating that single-token answer entropy effectively predicts answer correctness in knowledge-dependent domains (ROC-AUC up to 0.83) while revealing that entropy primarily captures knowledge uncertainty rather than reasoning difficulty. This work was recognized with the \textbf{Best Paper Award at AINL 2025}.

We discover and characterize language-specific directions in LLM latent space that can be identified through PCA with 96--100\% classification accuracy across four language pairs. We demonstrate that projecting out the source language component from hidden states reduces code-switching by 63--95\% in generated text without any additional training, providing a new mechanism for behavioral control.

We develop a complexity-aware fine-tuning pipeline that combines entropy-based data stratification with targeted training interventions---direct SFT for regular examples and chain-of-thought distillation for hard examples---achieving accuracy improvements while reducing training data requirements by 79--82\%, validated across three datasets and three model architectures.

Finally, we establish a unifying framework showing how signals from the forward pass, whether from output distributions or internal representations, can address diverse challenges in LLM deployment, suggesting broader applicability of these techniques.

\paragraph{Statements for defense.}

\begin{enumerate}
  \item Single-token answer entropy provides an effective signal for predicting LLM answer correctness in knowledge-dependent domains, achieving ROC-AUC scores up to 0.83 on the MMLU-Pro benchmark, while Model-as-Judge approaches perform near-randomly (ROC-AUC $\approx$ 0.49--0.57).

  \item Language-specific directions exist in the latent space of multilingual LLMs and can be reliably identified through PCA with 96--100\% classification accuracy across four language pairs. Projecting out the source language component from hidden states reduces code-switching by 63--95\% as measured by Code-Switching Index in generated text.

  \item Complexity-aware fine-tuning that stratifies training data by entropy and applies appropriate training strategies (direct SFT for regular examples, chain-of-thought distillation for hard ones) improves model accuracy while requiring 79--82\% less training data than full distillation, generalizing across three datasets and three model architectures.

  \item Signals from a single forward pass through an LLM---both output distributions and internal representations---encode sufficient information to address fundamental challenges in reliability, control, and training efficiency without requiring architectural modifications.
\end{enumerate}
