\chapter{Literature review}

This chapter surveys the research landscape relevant to the three main themes of this thesis: uncertainty estimation in language models, interpretability and control of internal representations, and efficient training methodologies. For each area, we review the current state of the art, identify key limitations, and position our contributions within the broader research context.

\section{Uncertainty Estimation in Large Language Models}

The problem of uncertainty quantification in neural networks has a long history~\cite{gal2016dropout}, but the advent of large language models has introduced new challenges and opportunities. Traditional approaches to uncertainty estimation, such as Monte Carlo dropout~\cite{gal2016dropout} and deep ensembles~\cite{lakshminarayanan2017simple}, require multiple forward passes through the model and are computationally prohibitive for models with billions of parameters. A single forward pass through a 70B-parameter model already requires significant computational resources; running it dozens or hundreds of times for ensemble-based uncertainty estimation is impractical in real-time applications.

This computational constraint has motivated research into more efficient alternatives. Confidence calibration methods attempt to align model confidence scores with actual accuracy~\cite{guo2017calibration}, but LLMs often exhibit poor calibration, particularly on out-of-distribution inputs~\cite{jiang2021can}. The gap between a model's apparent confidence and its actual correctness---often called the confidence gap---poses critical risks in high-stakes applications where overconfidence in incorrect answers leads to severe consequences. In domains such as healthcare, law, education, and economics, an LLM's overestimation of its accuracy could result in harmful decisions or misinformation.

Verbalized confidence, where models are prompted to express uncertainty in natural language~\cite{kadavath2022language,lin2022teaching}, offers an interesting alternative but relies on the model's ability to accurately self-assess---a capability that is itself uncertain and may not generalize across domains. Token-level probability analysis has emerged as a more principled direction. Kuhn et al.~\cite{kuhn2023semantic} introduced semantic entropy, which clusters generated sequences by meaning before computing entropy, addressing the challenge of semantically equivalent but lexically different outputs. Malinin and Gales~\cite{malinin2021uncertainty} proposed using the entropy of the predictive distribution as an uncertainty measure for sequence generation tasks. Fadeeva et al.~\cite{fadeeva2023lmpolygraph} provided a comprehensive comparison of white-box and black-box uncertainty estimation methods, presenting promising results for entropy-based aggregation approaches, while their subsequent work~\cite{fadeeva2024factchecking} demonstrated the potential of token-level uncertainty for fact-checking LLM outputs.

However, existing methods tend to answer how good an overall confidence estimate is while ignoring why uncertainty appeared in a specific case and how it relates to the nature of the question. Furthermore, many existing methods generalize poorly for concise multiple-choice questions where the model's response consists of only one or two tokens. The relationship between entropy patterns during generation and answer correctness, particularly how this relationship varies across question topics and reasoning requirements, remains underexplored. Our research addresses this gap by systematically analyzing how single-token answer entropy correlates with model reliability across different domains and complexity levels.

\section{Internal Representations and Interpretability}

Understanding the internal representations of neural networks has been a central goal of interpretability research~\cite{olah2017feature,elhage2022toy}. For language models, this involves understanding what information is encoded in activation patterns and how this information is processed across layers.

Probing classifiers have been widely used to identify what linguistic and semantic information is encoded in model representations~\cite{belinkov2017neural,conneau2018you}. These studies have revealed that different layers capture different types of information, with lower layers encoding syntactic features and higher layers encoding more semantic content~\cite{jawahar2019does}. This layered organization suggests that models develop increasingly abstract representations as information flows through the transformer stack.

More recently, researchers have discovered that model representations exhibit meaningful geometric structure beyond what probing classifiers capture. Linear directions in activation space have been found to correspond to interpretable concepts~\cite{mikolov2013linguistic,park2023linear}, leading to the ``linear representation hypothesis''---the idea that many high-level concepts are represented as directions in the model's activation space. This insight has enabled activation steering approaches, where model behavior can be modified by adding or removing vectors from intermediate activations~\cite{turner2023activation,li2024inference}. Zou et al.~\cite{zou2023representation} introduced representation engineering, a systematic framework that uses PCA on contrastive prompt pairs to identify concept directions (e.g., honesty, harmlessness) and steers activations along them to control model behavior.

In parallel with steering methods, concept erasure approaches have been developed to remove specific information from representations entirely. Iterative null-space projection (INLP)~\cite{ravfogel2020null} removes protected attributes by projecting representations onto the null space of a linear classifier trained to predict those attributes. LEACE~\cite{belrose2023leace} achieves the same goal in closed form, providing perfect linear concept erasure without iterative optimization. These methods demonstrate that specific properties encoded in representations can be surgically removed, a finding that is closely related to our approach of removing language identity via projection.

For multilingual models, the question of how different languages are represented has received significant attention. Work on mBERT and XLM-R revealed that language identity concentrates in a few principal components of hidden states, separable from semantics when analyzed on parallel texts~\cite{conneau2020emerging,chi2020finding}. Yang et al.~\cite{yang2021simple} demonstrated that removing these language-specific components produces more language-agnostic embeddings, improving cross-lingual transfer. Wendler et al.~\cite{wendler2024llamas} extended these findings to decoder-only LLMs, revealing that modern autoregressive models maintain interpretable geometric structure in their representations, with latent language preferences that can be identified through simple linear analysis. While these prior works focused primarily on analysis and understanding, our work takes the next step by exploiting this geometric structure for active generation control, using language identity as a practical test case for inference-time behavioral modification.

\section{Code-Switching in Multilingual Models}

Code-switching refers to the alternation between languages within a single conversation or text. While natural in multilingual human communication, unintended code-switching in LLM outputs is generally undesirable as it reduces output quality and user experience~\cite{etxaniz2024multilingual}. In deployed systems, such behavior degrades user trust: a customer-service chatbot prompted in Spanish that switches to English mid-response creates confusion rather than value. Recent work has also identified code-switching as a potential red-teaming vector, as language switches can bypass safety filters designed for specific languages.

Prior approaches to controlling code-switching have primarily relied on training-time interventions. Language-specific fine-tuning can reduce code-switching but requires separate models or adapters for each language~\cite{pfeiffer2020adapterhub}, which is expensive to maintain and does not scale to the hundreds of languages that modern LLMs nominally support. Constrained decoding methods can force outputs to remain in a target language but may degrade generation quality~\cite{post2018fast} by restricting the model's vocabulary in ways that interfere with fluent expression.

The discovery that language identity is encoded as linear directions in latent space~\cite{wendler2024llamas} opens new possibilities for inference-time control that avoid the costs of retraining. However, practical methods for exploiting this structure to reduce code-switching have not been thoroughly explored. Our work addresses this gap by developing and evaluating activation steering methods specifically designed for language control, demonstrating that a simple projection operation applied to the final layers can substantially reduce code-switching without any additional training.

\section{Efficient Fine-Tuning and Data Selection}

Fine-tuning large language models is computationally expensive, motivating research into more efficient approaches at multiple levels. Parameter-efficient fine-tuning methods such as LoRA~\cite{hu2022lora} and adapters~\cite{houlsby2019parameter} reduce the number of trainable parameters, making fine-tuning feasible on consumer hardware. However, these methods do not address data efficiency---they still train on all available data regardless of its informativeness.

Data selection and curriculum learning aim to improve training efficiency by carefully choosing which examples to train on and in what order. Influence functions~\cite{koh2017understanding} can identify the most impactful training examples but are prohibitively expensive to compute for models with billions of parameters. Active learning approaches~\cite{settles2009active} select informative examples iteratively but require multiple training rounds, each potentially as expensive as full fine-tuning. The LIMA approach~\cite{zhou2023lima} demonstrated that as few as 1,000 carefully curated high-quality examples suffice for alignment, showing that data quality and curation can substitute for data volume---but manual curation does not scale.

Recent work has explored using model predictions to guide data selection automatically. Self-training and self-distillation methods~\cite{furlanello2018born,xie2020self} use model outputs as targets for further training. Chain-of-thought distillation~\cite{ho2023large,magister2023teaching,hsieh2023distilling} transfers reasoning capabilities from larger to smaller models by training on generated rationales, achieving strong performance with smaller model sizes. The SmallToLarge approach~\cite{kim2024strategic} uses small models to summarize training trajectories and guide data selection for larger models, demonstrating that cheap proxy signals can inform expensive training decisions.

The idea of stratifying training data by difficulty has been explored in curriculum learning~\cite{bengio2009curriculum,kim2024strategic}, where models are trained on progressively harder examples. However, curriculum-style approaches have shown limited gains in final accuracy when applied to LLM fine-tuning, with extreme sensitivity to hyperparameters such as epoch counts and learning rates. Our work proposes a fundamentally different approach: rather than merely ordering examples by difficulty, we apply different training interventions to different complexity levels. Easy examples receive standard supervised fine-tuning, while hard examples receive chain-of-thought distillation. This insight---that different complexity levels benefit from different training strategies, not just different ordering---is the key contribution that distinguishes our approach from prior curriculum learning methods.

\section{Summary and Research Gaps}

The literature reveals several research gaps that this thesis addresses. While uncertainty estimation methods exist, the relationship between single-token answer entropy and answer correctness has not been systematically studied across different question domains and reasoning requirements, leaving practitioners without clear guidance on when to trust entropy-based confidence signals. Although language-specific directions in latent space have been identified in prior analytical work, practical methods for using them to control code-switching at inference time---particularly methods that require no additional training---are lacking. Finally, existing data selection methods do not leverage the insight that examples of different complexity may benefit from fundamentally different training strategies rather than merely different ordering or weighting.

This thesis addresses these gaps through three interconnected studies that exploit signals from the model's forward pass to improve reliability, control, and efficiency.
