\chapter{Discussion and conclusion}

\section{Summary of Contributions}

This thesis investigated how signals from a single forward pass through large language models---both output distributions and internal representations---can be leveraged to address fundamental challenges in LLM deployment. We presented three interconnected studies that demonstrate the practical utility of this approach.

In the first study, we demonstrated that single-token answer entropy provides an effective and practical signal for predicting answer correctness in knowledge-dependent domains. Across four models spanning parameter counts from 1.5B to 72B, we observed ROC-AUC scores up to 0.83 on the MMLU-Pro benchmark, with a clear scaling trend where larger models produce more informative entropy signals. A key finding is that entropy primarily captures knowledge uncertainty rather than reasoning difficulty: the signal is strongest for factual recall domains such as Biology and Economics, and weakest for reasoning-intensive domains such as Math and Law. The Model-as-Judge approach, by contrast, performs near-randomly (ROC-AUC $\approx$ 0.49), indicating that external evaluation captures fundamentally different properties than the model's own internal uncertainty. Calibration analysis revealed systematic overconfidence across all architectures, particularly in high-certainty regimes, suggesting that entropy-based monitoring could serve as a valuable safeguard in deployment. This work was recognized with the Best Paper Award at AINL 2025.

In the second study, we discovered that language-specific directions can be reliably identified in LLM latent space through PCA, achieving 96--100\% classification accuracy across four language pairs on Qwen2.5-1.5B and Llama-3.2-1B. The layer-wise analysis revealed that language identity crystallizes in final layers while early layers focus on semantic processing, suggesting these properties occupy largely orthogonal subspaces. By projecting out the source language component from hidden states during inference, we reduced distributional divergence by 25--55\% and achieved 63--95\% reduction in Code-Switching Index in generated text, without requiring any additional training. The method demonstrates that behavioral control can be achieved through geometric manipulation of internal representations, though the fragility of steering coefficients---where small changes in magnitude can shift output from fluent text to degenerate repetition---represents an important practical constraint that requires careful per-language calibration.

In the third study, we developed a training pipeline that uses single-token answer entropy to stratify training data by complexity and applies targeted interventions: direct supervised fine-tuning for regular examples and chain-of-thought distillation from a teacher model ensemble for hard examples. This approach achieved accuracy improvements from 0.39 to 0.52 (Qwen-3B) and 0.51 to 0.64 (Phi-4-Mini) on MMLU-Pro while using 79--82\% less training data than full distillation. The method generalizes across three datasets (MMLU-Pro, GSM8K, MedMCQA) and three model architectures, with particularly strong gains on mathematical reasoning (GSM8K: 0.82 vs. 0.79 for full distillation). The sensitivity analysis confirmed that single-token entropy substantially outperforms both MASJ-based metrics and chain-of-thought entropy variants for complexity estimation, and that standard curriculum learning is fragile and insufficient without differentiated training strategies.

\section{Unifying Insights}

The three studies share a common finding: the signals produced during a single forward pass through an LLM encode rich information that extends well beyond the immediate task of next-token prediction. Output distributions reflect model confidence in a domain-dependent manner, with high entropy indicating uncertainty and the relationship between entropy and correctness being strongest for knowledge-intensive questions where the model either possesses relevant training data or does not. Internal representations encode interpretable geometric structure, with language identity represented as linear directions in activation space that emerge progressively through layers and concentrate in the final transformer blocks. The model's uncertainty about a training example, as measured by output entropy, reflects the example's intrinsic difficulty in a way that is actionable for training---different complexity levels genuinely benefit from different training interventions rather than merely different ordering.

These insights suggest that LLMs develop internal representations of their own capabilities and limitations that can be accessed and exploited without explicit training for self-assessment. The forward pass produces, as a byproduct of next-token prediction, signals about confidence, language identity, and input complexity that are linearly accessible and practically useful. This observation has implications beyond the three specific applications explored here: any property that is linearly encoded in the model's representations is potentially amenable to similar extraction and manipulation techniques.

A particularly interesting connection emerges between Studies 1 and 3. Both exploit the same signal---output entropy---but for different purposes. In Study 1, entropy serves as a runtime indicator of answer reliability, enabling users to decide whether to trust a given response. In Study 3, the same entropy signal is used offline during training to allocate computational resources efficiently. This dual use of a single signal demonstrates the versatility of forward-pass information and suggests that a unified monitoring and training framework could leverage entropy for both deployment-time quality assurance and training-time data curation simultaneously.

A deeper unifying theme connects all three studies: at their core, they all explore uncertainty and the various ways it can be measured and exploited. The first study measures uncertainty directly through output entropy. The second identifies geometric structure in the latent space that separates language identity from semantic content. The third uses uncertainty as a proxy for example complexity to guide training. What ties these together is the prospect of \textit{uncertainty decomposition}. The language directions identified in Study 2 define an orthogonal decomposition of the representation space into a language-specific subspace and a language-agnostic complement. If uncertainty signals from Study 1 were analyzed along this decomposition, one could potentially disentangle language-specific uncertainty (e.g., the model is unsure which language to generate in) from language-agnostic uncertainty (e.g., the model genuinely lacks knowledge about the topic). Such a decomposition would provide more actionable uncertainty estimates: language-specific uncertainty could be resolved by steering, while language-agnostic uncertainty would signal a genuine knowledge gap requiring external intervention. This perspective suggests that the methods developed in this thesis are not three independent techniques but rather complementary components of a broader framework for structured uncertainty analysis in LLMs.

\section{Position in the Research Landscape}

Our work contributes to several active research areas. In the domain of uncertainty quantification, we extend prior work on semantic entropy~\cite{kuhn2023semantic} and confidence calibration~\cite{guo2017calibration} by demonstrating that even the simplest entropy measure---applied to a single answer token---provides strong predictive signal when analyzed at the appropriate granularity (per domain, per reasoning type). Our approach requires only a single forward pass, making it practical for real-time applications where ensemble methods or Monte Carlo dropout would be prohibitively expensive.

In interpretability and steering, we build on discoveries about linear structure in LLM representations~\cite{park2023linear,turner2023activation} and extend the concept erasure framework~\cite{ravfogel2020null,belrose2023leace} to active generation control. While prior work in representation engineering~\cite{zou2023representation} focused on concepts such as honesty and harmlessness, we demonstrate that language identity provides a particularly clean test case for activation steering, with near-perfect linear separability and clear practical motivation. Our finding that optimal steering requires active counter-steering beyond mere projection removal ($|s| > 1$) adds nuance to the linear representation hypothesis.

In efficient training, our complexity-aware approach relates to curriculum learning~\cite{bengio2009curriculum} and data selection~\cite{settles2009active}, but differs in a critical way: rather than simply ordering examples by difficulty, we apply fundamentally different training procedures to different complexity tiers. The insight that letting the model focus on what it can already solve directly, while giving extra guidance only where it struggles, yields a better allocation of limited model capacity than either uniform SFT or uniform distillation.

\section{Limitations}

Several limitations should be acknowledged, organized by the study they primarily affect.

Regarding model scale and diversity, the uncertainty estimation study uses models from 1.5B to 72B parameters, but the language steering and fine-tuning experiments are limited to small models in the 1--3B range. Whether the geometric properties of language directions and the effectiveness of entropy-based stratification transfer to models with 70B+ parameters or Mixture-of-Experts architectures remains an open question. The strong scaling trend observed in Study 1 suggests that larger models may produce even more useful signals, but this requires empirical validation.

Regarding language and task coverage, while language steering was validated on four language pairs spanning different typological distances (English--Chinese, English--Spanish, English--Russian, English--Hindi), low-resource languages, right-to-left scripts, and agglutinative or tonal languages remain unexplored. English--Hindi consistently showed the weakest improvements across all metrics, suggesting that typologically distant pairs with different scripts present additional challenges. On the task side, all three studies focus on question-answering settings; the applicability of entropy-based uncertainty estimation and complexity-aware fine-tuning to open-ended generation, coding tasks, or dialogue systems needs investigation.

The steering fragility observed in Study 2 represents a practical limitation. Optimal coefficients require manual grid search per language pair and depend on whether the optimization target is distributional similarity or generation quality. Small changes in coefficient magnitude can cause transitions from fluent output to degenerate repetition, and no principled method for automatic coefficient selection has been developed. The artificial code-switching evaluation setup, where the second half of each sentence is machine-translated, may not capture the full complexity of naturally occurring code-switching patterns at morphosyntactic boundaries.

For the complexity-aware fine-tuning pipeline, low entropy can still correspond to hallucinations, leading to imperfect identification of question complexity. The three-way equal split into easy, medium, and hard tiers was chosen for simplicity and may not be optimal for all datasets. In settings where a teacher model is unavailable or provides low-quality reasoning chains, the benefit of chain-of-thought distillation for hard examples would be diminished.

\section{Future Directions}

This research opens several promising directions for future investigation. The three types of signals explored in this thesis---entropy patterns, geometric structure of representations, and complexity scores---could be combined into unified frameworks for model monitoring and control. A deployment system that simultaneously monitors answer confidence via entropy, enforces language consistency via steering, and continuously improves via complexity-aware fine-tuning would represent a practical step toward more reliable LLM applications.

The methods could be extended to enable online adaptation during deployment. Rather than computing complexity scores offline on a fixed dataset, entropy could be monitored in real time to identify examples where the model is struggling, with automatic escalation to a larger model or human review. Similarly, language steering coefficients could be adapted dynamically based on observed code-switching patterns in production traffic.

Beyond language identity, the geometric structure of latent space could be exploited for steering other attributes such as formality, domain expertise, safety compliance, or reasoning style. The near-perfect linear separability we observed for language suggests that other high-level attributes may also be linearly encoded, making multi-dimensional steering a realistic prospect. Developing theoretical foundations for understanding why these signals are predictive---connecting entropy to information-theoretic notions of model knowledge and linear directions to learned feature representations---would strengthen the basis for future applications and help predict when these methods will succeed or fail.

A particularly promising direction is structured uncertainty decomposition. The language directions identified in Study 2 define orthogonal subspaces in the representation space; projecting entropy signals from Study 1 onto these subspaces could separate language-specific uncertainty from language-agnostic uncertainty. Such decomposition would make uncertainty estimates more actionable: language-specific uncertainty could be resolved cheaply via steering, while language-agnostic uncertainty would signal a genuine knowledge gap requiring external intervention or abstention.

Finally, integrating the uncertainty and control mechanisms developed here with reinforcement learning from human feedback and other alignment techniques could improve model safety. Entropy-based confidence scores could modulate the strength of safety interventions, applying stronger guardrails when the model is uncertain and allowing more freedom when it is confident. This adaptive approach could reduce the tension between helpfulness and safety that characterizes current alignment methods.

\section{Conclusion}

This thesis demonstrated that accessible signals from a single forward pass through large language models provide valuable information for addressing fundamental deployment challenges. Single-token answer entropy predicts answer correctness with ROC-AUC up to 0.83 in knowledge-dependent domains, latent space geometry enables language control with 63--95\% reduction in code-switching, and complexity-aware data stratification improves training efficiency while using 79--82\% less data. Together, these contributions establish that exploiting forward-pass signals offers a practical path toward more reliable, controllable, and efficient LLMs---without requiring architectural modifications or expensive retraining.

\newpage
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to express my gratitude to my supervisor, Alexey Zaytsev, for his guidance and support throughout this research. I also thank my co-authors---Petr Sychev, Daniil Vyazhev, Edvard Khalafyan, and Nikolai Kondusov---for their valuable contributions to the published papers that form the basis of this thesis.

This research was conducted at Skolkovo Institute of Science and Technology (Skoltech). I acknowledge the computational resources provided by the Skoltech HPC cluster. The research was partially supported by the Russian Science Foundation grant No. 25-11-00355.
